Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-28 paper/paper.tex
latest - 28th

The  semantics  of   expressions  (and  applications)
%% \footnote{In  most
%%   contexts, we  shall use the  term 'expression' and the  notation $e$
%%   for both expressions and applications.} 
are given by transitions of
the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
Here \stk\ is a stack of continuation frames.  Each continuation frame
is of the form $(\rho, \ell,  e)$, signifying that the location $\ell$
has  to  be  updated  with  the  value  of  the  currently  evaluating
expression, and $e$ is to be evaluated next in the environment $\rho$.
Note  that  the  state  of  the  transition  system  consists  of  the
\emph{current evaluation  context}, given by  $\rho$ and $e$,  and the
\emph{suspended  evaluation  contexts}  (represented  as  continuation
frames) in the stack $S$.   
%% \warning{The
%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
%% system will be denoted $\mathcal{S}$.}

@@

\newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}

\subsection{Generating Equations for \Lfonly$_\mathit{f}$}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{1}{x_i}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$ at
the program point at  the beginning of $e_{\mathit{f}}$
(assumed   to   be   $\pi_1$).   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:

\begin{eqnarray*}
&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}

%% and  the only equation defining \Lfone{\length} is:
%%   \begin{eqnarray*}
%%    && \Lf{\length}{1}{\sigma}
%%     =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
%% \clazy\Lf{\length}{1}{\clazy\sigma}
%% \end{eqnarray*}
and  the equation defining \Lfone{\fun} is:
\begin{eqnarray*}
  && \Lf{\fun}{1}{\sigma}
  =  \clazy\sigma \cup \Lf{\length}{1}{\bcar\sigma_{\mathit{\length}}} \cup  \Lf{\length}{1}{\clazy\sigma} 
  \cup  \acdr\Lf{\fun}{1}{\bcdr\sigma_{\mathit{\fun}}}
\end{eqnarray*}

In general, the equations for \Lfonly\ are recursive as
in the  case of  \Lfonly$_\mathit{f}$. As  mentioned in
\cite{asati14lgc},   a   closed   form   solution   for
$\Lfone{\mathit{f}}$ can  be derived by  observing that
each of  the liveness rules  modifies a demand  only by
prefixing  it with  symbols  in  the alphabet  $\lbrace
\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore
we can  assume that $\Lf{f}{i}{\sigma}$ has  the closed
form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}

where \Df{f}{i}  are sets of strings  over the alphabet
mentioned above.  Substituting the  guessed form in the
equation   describing    \Lfonly$_{\mathit   f}$,   and
factoring  out   $\sigma$,  we  get  an   equation  for
\Df{f}{i}  that   is  independent  of   $\sigma$.   Any
solution   for   \Df{f}{i}   yields  a   solution   for
\Lfonly$_{\mathit  f}$.   Applied to  \Lfonly$_{\mathit
  \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\fun }{1}{\sigma} = \Df{\fun}{1}\sigma,~\mbox{and}\\
&&   \Df{\fun}{1} =  \clazy \cup \Df{\length}{1}\bcar \cup  \Df{\length}{1}\clazy 
       \cup  \acdr\Df{\fun}{1}\bcdr
  \end{eqnarray*}

Note that this equation can also be viewed as a CFG with \{\acdr, \bcar, \bcdr,
\clazy\} as terminal symbols and \Df{\length}{1} and \Df{\fun}{1} as the 
non-terminals.

@@

For  the  running  example,   $\fun$  has  calls  from  $\main$  at
$\pi_{16}$ and a recursive call at $\pi_6$.
So $\sigma_{\fun} =
     \deltacall{\fun}{\pi_{16}}{\main}  \cup
\deltacall{\fun}{\pi_6}{\fun}$.

%% For  the  running  example,   $\length$  has  calls  from  $\main$  at
%% $\pi_{12}$ and a recursive call at $\pi_6$.
%% So $\sigma_{\length} =
%%      \deltacall{\length}{\pi_{12}}{\main}  \cup
%% \deltacall{\length}{\pi_6}{\length}$.

Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\fun}    &=&
 \{\sigma_{\mathit{all}}\}  ~\cup~{\bcdr\sigma_{\mathit{\fun}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{2}{\pl} and \Lanv{1}{\py} in terms
of  $\sigma_{\fun}$ are:
\begin{align*}
\Lanv{2}{\pl} &= \clazy \cup \acar\Lf{\fun}{2}{\bcdr\sigma_{\mathit{\fun}}}
 \\
\Lanv{1}{\py} &= \clazy \cup \bcar\sigma_{\mathit{\fun}}
\end{align*}

In  summary, the  equations  generated during  liveness
analysis are:
\begin{enumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$  for  $e_\mainpgm$) an  equation  defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}      The
equations  above   can  now  be  re-interpreted   as  a
context-free    grammar   (CFG)    on   the    alphabet
$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
Let \var{$X$}  denote the  non-terminal for  a variable
$X$  occurring on  the LHS  of the  equations generated
from  the  analysis.  We  can  think  of the  resulting
productions as being  associated with several grammars,
one for each non-terminal \var{\Lanv{i}{x}} regarded as
a  start  symbol.   As  an  example,  the  grammar  for
\var{\Lanv{1}{\xl}}     comprises     the     following
productions:
\begin{eqnarray*}
  \var{\Lanv{3}{\xl}}  &\rightarrow& 
  \clazy \mid \var{\Df {\Sum}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}\clazy \mid \acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\clazy  \mid \bcar\langle{\sigma_{\Sum}}\rangle \\
 \var{\Df{\fun}{2}} &\rightarrow& \var{\Df{\makelist}{1}}\clazy \mid \var{\Df{\makelist}{1}}\var{\Df{\Sum}{1}}\clazy \\ 
& & \mid \var{\Df{\makelist}{1}}\acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
\langle\sigma_{\mathit{\fun}} \rangle &\rightarrow& \sigma_{all} \mid \bcdr\var{\sigma_{\mathit{\fun}}} \\
\var{\Df{\Sum}{1}} &\rightarrow& \clazy \mid \acar\clazy \mid \acdr\var{\Df{\Sum}{1}}\clazy
%% \var{\Lanv{3}{\py}} &\rightarrow& \clazy \mid \bcar\var{\sigma_{\mathit{\fun}}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{i}{x}},   denoted
$\mathscr{L}(\var{\Lanv{i}{x}})$,   is    the   desired
solution  of  \Lanv{i}{x}.

\warning{
\hrule
\medskip
Fixing from here
\medskip
\hrule
}

However, recall that the decision problem that we are interested in
during garbage collection is the following:
\begin{quote}
Let $x.\alpha$ be a forward access path consisting only
of edges  \acar\ and \acdr\  (but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
\end{quote}

@@

%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
%% that the results of liveness analysis of any program can be modeled by
%% the kind  of grammar described above. The  following proposition shows
%% that the converse also holds.
%% \begin{proposition}
%% Given    a     grammar    $G$    of    the     form    described    in
%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
%% program $p$  with program points  $\pi_i$ and variables $x$  such that
%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
%% \end{proposition}
We first show that membership problem of the class of grammars in
Definition~\ref{def:specialgrammar} is undecidable. However note that
the grammar may not be necessarily generated from liveness analysis.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
\begin{lemma}\label{lemma:grammar-undecidable}
Consider a grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The membership problem
$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

@@

\begin{lemma}
Consider a grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} that is generated by liveness analysis
of a program and a forward access path $\alpha$. The membership problem
$\alpha\$ \in \mathscr{L}(G)$ is also undecidable.
\end{lemma}

Given a Turing machine and an input string, we now give program fragments whose liveness
analysis generate the same productions as in
Lemma~\ref{lemma:grammar-undecidable}, but with  extra productions
that necessarily contain the \clazy\ symbol. For instance,
corresponding to the production $\nont{i}{c} \rightarrow c
\nont{j}{\acar}\bcar$, with $c$ as \acar\ for concreteness, we have
the program (define ($\nont{i}{c}$ a) (let b <- (car a) in (let c <-
($\nont{j}{\acar}$ b) in (let d <- 5 in (let e <- (cons c d) in (let f
<- (null? e) in return f)))))). The reader can check that this
produces the grammar fragment ...

It is obvious that if the TM halts on the input, then the grammar will
produce the $\epsilon$ string. However, if the TM does not halt on the
input, then the productions containg

>>>>

\warning{The  semantics  of   expressions  (and  applications)
%% \footnote{In  most
%%   contexts, we  shall use the  term 'expression' and the  notation $e$
%%   for both expressions and applications.} 
are given by transitions of the form $\rho, \stk, \heap, e \rightarrow
\rho',  \stk', \heap',  e'$.  Here  \stk\ is  a stack  of continuation
frames.  Each  continuation frame  is of the  form $(\rho,  \ell, e)$,
signifying that the  location $\ell$ has to be updated  with the value
of the  currently evaluating  expression, and $e$  is to  be evaluated
next  in the  environment $\rho$.   We shall call the $\rho$ and
the $e$ on the LHS of a $ \stk,
\heap,  e$ is the state of  the  transition system, then we shall there is a   \emph{current
  evaluation  context}, given  by  $\rho$, $\heap$  and  $e$, and  
\emph{suspended  evaluation  contexts}  (represented  by the
environment, heap and expression in each   continuation
frame) in the stack $S$.}
%% \warning{The
%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
%% system will be denoted $\mathcal{S}$.}

@@

\subsection{Generating Equations for $\Lfonly_{\mathit f}$}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{1}{x_i}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$ at
the program point at  the beginning of $e_{\mathit{f}}$
(assumed   to   be   $\pi_1$).   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
$$
 \Lf{\length}{1}{\sigma} = \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
$$
%% and  the only equation defining \Lfone{\length} is:
%%   \begin{eqnarray*}
%%    && \Lf{\length}{1}{\sigma}
%%     =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
%% \clazy\Lf{\length}{1}{\clazy\sigma}
%% \end{eqnarray*}

In general, the equations for \Lfonly\ are recursive as
in the  case of  \Lfonly$_\mathit{f}$. As  mentioned in
\cite{asati14lgc},   a   closed   form   solution   for
$\Lfone{\mathit{f}}$ can  be derived by  observing that
each of  the liveness rules  modifies a demand  only by
prefixing  it with  symbols  in  the alphabet  $\lbrace
\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore
we can  assume that $\Lf{f}{i}{\sigma}$ has  the closed
form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i}  are sets of strings  over the alphabet
mentioned above.  Substituting the  guessed form in the
equation   describing    \Lfonly$_{\mathit   f}$,   and
factoring  out   $\sigma$,  we  get  an   equation  for
\Df{f}{i}  that   is  independent  of   $\sigma$.   Any
solution   for   \Df{f}{i}   yields  a   solution   for
\Lfonly$_{\mathit  f}$.   Applied to  \Lfonly$_{\mathit
  \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy
       \cup \clazy\Df{\length}{1}{\clazy}
  \end{eqnarray*}

Note that this equation can also be viewed as a CFG with \{\acdr,
\clazy\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.

@@

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{12}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{12}}{\main}  \cup
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
 \\
\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
\end{align*}

In  summary, the  equations  generated during  liveness
analysis are:
\begin{enumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$  for  $e_\mainpgm$) an  equation  defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}      The
equations  above   can  now  be  re-interpreted   as  a
context-free    grammar   (CFG)    on   the    alphabet
$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
Let \var{$X$}  denote the  non-terminal for  a variable
$X$  occurring on  the LHS  of the  equations generated
from  the  analysis.  We  can  think  of the  resulting
productions as being  associated with several grammars,
one for each non-terminal \var{\Lanv{i}{x}} regarded as
a  start  symbol.   As  an  example,  the  grammar  for
\var{\Lanv{1}{\xl}}     comprises     the     following
productions:
\begin{eqnarray*}
  \var{\Lanv{1}{\xl}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
\var{\Lanv{9}{\pa}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{i}{x}},   denoted
$\mathscr{L}(\var{\Lanv{i}{x}})$,   is    the   desired
solution  of  \Lanv{i}{x}.    
However, recall that the decision problem that we are interested in
during garbage collection is the following: 
\begin{quote}
Let $x.\alpha$ be a forward access path consisting only
of edges  \acar\ and \acdr\  (but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
\end{quote}

@@

%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
%% that the results of liveness analysis of any program can be modeled by
%% the kind  of grammar described above. The  following proposition shows
%% that the converse also holds.
%% \begin{proposition}
%% Given    a     grammar    $G$    of    the     form    described    in
%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
%% program $p$  with program points  $\pi_i$ and variables $x$  such that
%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
%% \end{proposition}
We first show that membership problem of the class of grammars in
Definition~\ref{def:specialgrammar} is undecidable. However note that
the grammars may not be necessarily generated from liveness analysis.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
\begin{lemma}\label{lemma:grammar-undecidable}
Consider a grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The membership problem
$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

@@

\begin{lemma}\ref{lemma:grammar-from-analysis-undecidable}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
analysis  of a  program,  and  a forward  access  path $\alpha$.   The
membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

\begin{proof}
Given a Turing machine and an input string, we shall define a function
for   each  of   the  non-terminals   introduced  in   the  proof   of
Lemma~\ref{lemma:grammar-undecidable}.
%% Liveness  analysis  of  $f$  will  result in  not  only  contain  a
%% production  $P$ itself,  but could  also contain  extra productions
%% that necessarily contain the \clazy\ symbol.
For instance, corresponding to the production $\nont{i}{c} \rightarrow
c \nont{j}{\acar}\bcar$, with $c$ as \acar\ for concreteness, consider
the program:

If there more than one productions with the same LHS, then the
corresponding function creates joins them with a $\SIF$.   If there
are $n$ such functions,  then \mainpgm\  creates  a $n$-way
branch  and inserts  a  single call  to a  distinct  function in  each
branch. It can  be verified that the relevant part  of the grammar for
the production $\nont{i}{c} \rightarrow
c \nont{j}{\acar}\bcar$ shown earlier is:

Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
had a single non-terminal on the RHS. There are two characteristics of
the grammar produced from liveness-analysis  that are are relevant for
replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
grammar  generated from  liveness  analysis of the constructed program
will  have the  production
$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
\nont{\mathit{all}}{}$,  and (b)  other  productions are  of the  form
$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
\nont{\mathit{all}}{}$, where $\beta_1'$ and $\beta_2'$ do not contain
any non-terminals.

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
replacing   $\nont{\mathit{all}}{}$   in   the  sentential   form   by
$\epsilon$. However, if the TM does  not accept the input string, then
any  sentential  form derived  from  the  start  symbol would  have  a
non-terminal different from $\nont{\mathit{final}}{c}$. Thus the empty
string would not be derivable from the grammar.
\end{proof}
